### 2024-04-17 paper/paper.qmd
Added data cleaning,and usage txt

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
library(here)
library(ggplot2)

merged_data <- read.csv(here("data/analysis_data/merged_data.csv"))

```

# Introduction

The Canadian housing market has undergone significant changes in recent years, driven by various economic factors. This analysis aims to explore the dynamics of the housing market and understand the impact of economic variables on its stability.

The housing market is a critical component of the economy, influencing both individual financial well-being and broader economic stability. Housing affordability, in particular, has become a pressing issue, with many Canadians finding it increasingly challenging to purchase or rent a home. Toronto’s benchmark home price has risen by 42.8 per cent since 2019. According to the Canadian Real Estate Association (CREA), the benchmark home price in Toronto five years ago was $746,500. Now, in 2024, that number is $1,065,800.[@ctvnews_home_prices].

The Bank of Canada increased interest rates across the country 10 times between March, 2022 and July, 2023, bringing its benchmark rate to 5 per cent from 0.25 per cent in one of the most aggressive monetary policy campaigns on record.

Inflation in early 2024 seems to be settling, but 2023 it was at a high of

With so many factors at play, it is hard to keep track of just how the housing market is evolving.

Understanding the relationship between economic variables such as inflation, interest rates, unemployment, and housing prices is essential for policymakers, economists, and individuals alike. These variables play a significant role in shaping the housing market, impacting affordability, availability, and overall market stability.

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Data Source and Collection 
The data sets used for this analysis have been picked from multiple sources, open government portals and economic institutions. By choosing a variety of data sets, many cleaning and alignment processes were applied to ensure as much consistency possible between the data sets before usage.

### Inflation Rate 
The Inflation Rate Data was found from Macrotrends which published a concise summary of the inflation rate (Consumer Price Index focused) changes from 1960-2023 sourcing its data from the World Bank. Inflation rate, from a Consumer Price Index (CPI) perspective, represents the percentage change in the average price level of a basket of goods and services typically consumed by households over a specific period. It indicates the rate at which the general level of prices for goods and services is rising, leading to a decrease in purchasing power.

### Interest Rate 
The interest rate used for our analysis is specifically the Canada Mortgage and Housing Corporation, conventional mortgage lending rate, 5-year term rate, and was sourced from Statistics Canada. ()
The conventional mortgage lending rate, specifically the 5-year term, refers to the interest rate that financial institutions charge for conventional mortgages with a fixed interest rate and a term of 5 years. This rate is significant in the housing market as it influences the cost of borrowing for home buyers and impacts the overall affordability of housing.

### Unemployment Rate

### Housing Price Index

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: int-inf-plot
#| fig-cap: Relationship between inflation rate, interest rate, and unemployment rate
#| echo: false
#| warning: false
#| message: false
# Create a line plot with 'Inflation Rate', 'Interest Rate', and 'Unemployment Rate' over the years
ggplot(merged_data, aes(x = Year)) +
  geom_line(aes(y = `Inflation.Rate....`, color = "Inflation Rate")) +
  geom_line(aes(y = Interest_Rate, color = "Interest Rate")) +
  geom_line(aes(y = `Unemployment.Rate....`, color = "Unemployment Rate")) +
  labs(title = "Inflation Rate, Interest Rate, and Unemployment Rate Over the Years",
       x = "Year",
       y = "Rate") +
  scale_color_manual(values = c("Inflation Rate" = "blue", "Interest Rate" = "red", "Unemployment Rate" = "green")) +
  theme_minimal()

```

```{r}
#| label: hpi-plot
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false
# Assuming 'merged_data' is your merged dataset
ggplot(merged_data, aes(x = Year, y = Average_House_Only)) +
  geom_line() +
  labs(title = "Housing Price Index Over Time",
       x = "Year",
       y = "Average House Price Index") +
  theme_minimal()

```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: immigration-plot
#| fig-cap: Number of immigrants in Canada from 2000 to 2023
#| echo: false
#| warning: false
#| message: false
# Create a line plot for the number of immigrants in Canada over time
ggplot(merged_data, aes(x = Year, y = `Number.of.immigrants.in.Canada.from.2000.to.2023`)) +
  geom_line(color = "blue") +
  labs(title = "Number of Immigrants in Canada from 2000 to 2023",
       x = "Year",
       y = "Number of Immigrants") +
  theme_minimal()

```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-04-17 scripts/00-simulate_data.R

@@ -1,19 +1,48 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulate Data 
+# Author: Navya Hooda
+# Date: April 9, 2023
+# Contact: navya.hooda@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
 
+# Set seed for reproducibility
+set.seed(123)
 
-#### Workspace setup ####
-library(tidyverse)
-# [...UPDATE THIS...]
+# Number of observations
+num_observations <- 100
 
-#### Simulate data ####
-# [...ADD CODE HERE...]
+# Simulate data for new housing added in the last few years
+new_housing_added <- rnorm(num_observations, mean = 500, sd = 100)
+
+# Simulate data for proposed housing units in the next 2-3 years
+proposed_housing_units <- rnorm(num_observations, mean = 300, sd = 50)
+
+# Simulate data for housing rates
+housing_rates <- 50000 + 10*new_housing_added - 5*proposed_housing_units + rnorm(num_observations, mean = 0, sd = 10000)
+
+# Create a data frame to store the simulated data
+simulated_data <- data.frame(
+  New_Housing_Added = new_housing_added,
+  Proposed_Housing_Units = proposed_housing_units,
+  Housing_Rates = housing_rates
+)
+
+# Display the first few rows of the simulated data
+head(simulated_data)
+
+# Create a scatter plot of New_Housing_Added against Housing_Rates
+plot(simulated_data$New_Housing_Added, simulated_data$Housing_Rates,
+     xlab = "New Housing Added", ylab = "Housing Rates",
+     main = "Scatter plot of New Housing Added vs Housing Rates",
+     col = "blue", pch = 16
+)
+
+# Create a scatter plot of Proposed_Housing_Units against Housing_Rates
+plot(simulated_data$Proposed_Housing_Units, simulated_data$Housing_Rates,
+     xlab = "Proposed Housing Units", ylab = "Housing Rates",
+     main = "Scatter plot of Proposed Housing Units vs Housing Rates",
+     col = "blue", pch = 16
+)
 
 
 

### 2024-04-17 scripts/02-data_cleaning.R

@@ -1,44 +1,265 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Clean data for meaningful analysis
+# Author: Navya Hooda
+# Date: April 9, 2023
+# Contact: navya.hooda@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+
 
 #### Workspace setup ####
 library(tidyverse)
+library(dplyr)
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+inflation_data <- read_csv("data/raw_data/inflation-rate-cpi.csv", skip = 16)
+unemployment_data <- read_csv("data/raw_data/canada-unemployment-rate.csv", skip = 16)
+interest_data <- read_excel("data/raw_data/interest_rate_2000.xlsx",skip=, col_types = "text")
+hpi_index <- read_excel("data/raw_data/HPI_20-23.xlsx",skip=, col_types = "text")
+immigration <- read_excel("data/raw_data/immigration_data.xlsx",skip=, col_types = "text")
+# inflation
+# Clean data
+inflation_data_cleaned <- inflation_data %>%
+  # Convert 'date' to date format and extract the year
+  mutate(year = as.integer(substring(date, 1, 4))) %>%
+  # Filter for years 2013-2023
+  filter(year >= 2000 & year <= 2023) %>%
+  # Select relevant columns
+  select(year, `Inflation Rate (%)`, `Annual Change`) %>%
+  # Replace NA values in 'Annual Change' with 0
+  mutate(`Annual Change` = ifelse(is.na(`Annual Change`), 0, `Annual Change`))
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
[…]

### 2024-04-18 paper/paper.qmd
updated

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "Economic Drivers, Immigration Dynamics, and Housing Starts: Their Collective Influence on Housing Price Index"
+subtitle: ""
 author: 
-  - First author
-  - Another author
-thanks: "Code and data are available at: LINK."
+  - Navya Hooda
+
+thanks: "Code and data are available at: https://github.com/hoodanav/Housing_Market_Analysis/tree/main"
 date: today
@@
 library(ggplot2)
+library(scales)
 
 merged_data <- read.csv(here("data/analysis_data/merged_data.csv"))
+housing_starts <- read.csv(here("data/analysis_data/total_starts_by_year.csv"))
 
@@
 
-The Canadian housing market has undergone significant changes in recent years, driven by various economic factors.
-This analysis aims to explore the dynamics of the housing market and understand the impact of economic variables on its stability.
+In recent years, the Canadian housing market has undergone significant changes due to various economic factors.
+Canadians are facing a housing crisis driven by high demand, unaffordable rates, and population growth in major cities across Canada.
+For example, Toronto's benchmark home price has surged by 42.8% since 2019.
+Back in 2019, the benchmark home price in Toronto stood at $746,500.
+Fast forward to 2024, and that figure has soared to $1,065,800, according to the Canadian Real Estate Association (CREA).
 
-The housing market is a critical component of the economy, influencing both individual financial well-being and broader economic stability.
-Housing affordability, in particular, has become a pressing issue, with many Canadians finding it increasingly challenging to purchase or rent a home.
-Toronto’s benchmark home price has risen by 42.8 per cent since 2019.
-According to the Canadian Real Estate Association (CREA), the benchmark home price in Toronto five years ago was $746,500.
-Now, in 2024, that number is $1,065,800.[@ctvnews_home_prices].
+One major factor contributing to these changes is the notable increase in immigration over the last decade.
+In 2023, Canada experienced its highest influx of immigrants and students in 70 years, leading to a 2.9% surge in the population [@reuters_canada_population].
+While achieving immigration targets is a milestone for Canada, the influx of newcomers has likely exacerbated the already tight housing market.
 
-The Bank of Canada increased interest rates across the country 10 times between March, 2022 and July, 2023, bringing its benchmark rate to 5 per cent from 0.25 per cent in one of the most aggressive monetary policy campaigns on record.
+Globally, economic shifts in recent years have had far-reaching effects.
+Inflation has risen not only in Canada but also in many countries worldwide due to various events.
+Supply chain disruptions on a global scale have driven prices higher, while geopolitical tensions, such as Russia's invasion of Ukraine, have further contributed to increased energy and food costs [@forbes_interest_rates].
 
-Inflation in early 2024 seems to be settling, but 2023 it was at a high of
+In response to rising inflation, Canada saw its inflation rate peak at 6.1% in 2022, gradually easing to 3% in 2023.
+However, the target rate for stability is around 2% [@boc_inflation].
+To curb inflation, the Bank of Canada implemented a series of interest rate hikes between March 2022 and July 2023.
+This aggressive monetary policy campaign saw the benchmark rate soar from 0.25% to 5%, marking one of the most substantial increases on record[@globeandmail_boc].
 
 With so many factors at play, it is hard to keep track of just how the housing market is evolving.
+This analysis aims to explore the dynamics of the housing market and understand the impact of economic variables on the housing market, analyze if any indicators are more significant contributors to the housing market than others etc.
+Through this exploration we also aim to explore economic indicator changes and what significance they have to the housing market, the relationship between correlated variables, and their impact on the market.
 
-Understanding the relationship between economic variables such as inflation, interest rates, unemployment, and housing prices is essential for policymakers, economists, and individuals alike.
-These variables play a significant role in shaping the housing market, impacting affordability, availability, and overall market stability.
+This paper also explores the housing supply in specifically Ontario due to its increased housing activity in the past few years.
+We explore how the housing demand has changed over the last 20 years, and whether the current housing development initiatives are enough to contribute positively to the improvement of the housing crisis.
+Additionally, we touch upon government policy on housing capacity, and targets that are currently in motion.
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+With the complexity of the housing market influenced by numerous factors, keeping track of its evolution proves challenging.
+This analysis aims to understand the dynamics of the housing market, exploring the impact of economic variables and identifying significant contributors.
+By examining economic indicator changes, we seek to uncover their significance and understand correlations among variables and their market impact.
 
-The remainder of this paper is structured as follows. @sec-data....
+Focusing particularly on Ontario's housing landscape, where activity has surged in recent years, this paper investigates shifts in housing demand over the past two decades.
+We evaluate whether current housing development initiatives adequately address the housing crisis.
+Additionally, we discuss government policies regarding housing capacity and existing targets.
+
+Through this analysis, we aim to grasp the current reality of the housing market and understand how economic conditions, including factors like immigration and housing starts, influence its trajectory.
+With the housing market facing constraints such as high interest rates, we seek to uncover insights regarding the future of the market.
+
+The remainder of this paper is structured as follows.In the Data@sec-data, we note the sources of the data sets utilized in the paper, the specific techniques we used to process them, and key variables we wish to use.
+In the Model @sec-data2, we created models to assess the impact of the key variables on our outcome of housing prices.
+In the Results @sec-data3 we discuss  the results of the model and what key correlations were found from the variables studied in the paper.
+In the Discussion @sec-data4, we explore the findings of modeling and predictions.
+In addition, we discuss biases and weaknesses in the data that contributed to our findings, and how we navigated those limitations.
 
@@
 ### Inflation Rate 
-The Inflation Rate Data was found from Macrotrends which published a concise summary of the inflation rate (Consumer Price Index focused) changes from 1960-2023 sourcing its data from the World Bank.
+The Inflation Rate Data was found from Macrotrends [@inflation] which published a concise summary of the inflation rate (Consumer Price Index focused) changes from 1960-2023 sourcing its data from the World Bank.
 Inflation rate, from a Consumer Price Index (CPI) perspective, represents the percentage change in the average price level of a basket of goods and services typically consumed by households over a specific period.
@@
 ### Interest Rate 
-The interest rate used for our analysis is specifically the Canada Mortgage and Housing Corporation, conventional mortgage lending rate, 5-year term rate, and was sourced from Statistics Canada. ()
+The interest rate used for our analysis is specifically the Canada Mortgage and Housing Corporation, conventional mortgage lending rate, 5-year term rate, and was sourced from Statistics Canada [@interest].
+The data was reported for years 2000-2024 by each month.
 The conventional mortgage lending rate, specifically the 5-year term, refers to the interest rate that financial institutions charge for conventional mortgages with a fixed interest rate and a term of 5 years.
 This rate is significant in the housing market as it influences the cost of borrowing for home buyers and impacts the overall affordability of housing.
+For our purposes we chose to aggregate data to use an average over all months to gain a yearly interest rate number to explore.
 
 ### Unemployment Rate
+The unemployment rate was found from Macrotrends [@unemployment_rate] which gave summarized year round rates, initially sourced from the World Bank.
+The unemployment rate generally refers to the percentage of the labor force that is unemployed and actively seeking employment.
+A visual of the rates discussed so far is seen in @fig-int-inf-plot.
 
-### Housing Price Index
+### New Housing Price Index (HPI)
+The  new housing price index data was found from Statistics Canada [@hpi], and was reported by each month for each year from 2000-2024 specifically covering Ontario.
+The new housing price index is a value that is a monthly series that measures changes over time in the contractors' selling prices of new residential houses given a specific start date (typically with HPI = 100, currently this refers to December 2016).
+The data is reported monthly on NHPI, a survey that collects the following dwelling types: new single homes, semi-detached homes and town homes (row or garden homes).
+The survey also collects builders' estimates of the current value (evaluated at market price) of the land.
+The data reports HPI for land type, house type, and total type.
+For our purposes we chose the house type parameter and sought to calculate a yearly HPI given the data was sorted by months.
+The trend can be seen in (@fig-hpi-plot)
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+### Immigration Numbers
+The yearly immigration numbers were obtained from Statista [@immigrants], and was originally sourced from Statistics Canada.
+This data reports the number of immigrants Canada welcomes in each year from 2000-2023.
+The data was collected and sourced from a survey done by Statistics Canada.
+A visual of these numbers is seen in @fig-immigration-plot.
+
+### Housing Starts (Ontario)
+The housing starts data was found from Statistics Canada [@housing_starts], was reported quarterly from years 2000-2024.
+This data refers to the number of units of housing started across Ontario of different dwelling types, in various regions.
+We chose to look at the housing starts of total units for all dwelling types, a yearly number by summing up each quarter's progress.
+For the 2024 value only quarter 1 numbers are reported hence the sharp decline for 2024 on @fig-housing-plot
+
+## Data Cleaning 
+Simple data cleaning operations were applied to achieve a cohesive merged data set for analysis.
+We used R [@citeR] for data cleaning and processing, using packages like tidyverse [@tidy] for data manipulation and janitor [@jan] for cleaning column names.
+Other packages used includes `ggplot2` [@ggplot], `dplyr` [@dp], `readr` [@read], `tibble` [@tib], `janitor` [@jan], `knitr` [@knit], `ggbeeswarm` [@gg], `ggrepel` [@repel], `kableExtra`[@kable], `readxl`[@readxl], `rstanarm`[@rstan], `modelsummary`[@model] and `here` [@here].
+All these different data sets were cleaned in a different way to match the final representation of the cleaned data.
 
 ```{r}
-#| label: int-inf-plot
-#| fig-cap: Relationship between inflation rate, interest rate, and unemployment rate
+#| label: fig-int-inf-plot
 #| echo: false
 #| warning: false
+#| fig-cap: Inflation, Interest and Unemployment Rates 2000-2023
 #| message: false
@@
   geom_line(aes(y = `Unemployment.Rate....`, color = "Unemployment Rate")) +
-  labs(title = "Inflation Rate, Interest Rate, and Unemployment Rate Over the Years",
+  labs(title = ,
        x = "Year",
@@
 
+## Data Modifications
+Overall, the values were often reported at different times, for consistency we chose to use a yearly metric. We averaged upon quarterly or monthly reported data to gain an averaged yearly value.
+
 ```{r}
-#| label: hpi-plot
-#| fig-cap: Relationship between wing length and width
+# Load the necessary packages
+#| label: fig-immigration-plot
+#| fig-cap: Number of immigrants in Canada from 2000 to 2023
 #| echo: false
+#| warning: false
+#| message: false
+# Filter out rows with missing values in the 'Number.of.immigrants.in.Canada.from.2000.to.2023' column
+merged_data_filtered <- merged_data[!is.na(merged_data$Number.of.immigrants.in.Canada.from.2000.to.2023),]
+
+# Create a line plot for the number of immigrants in Canada over time
+ggplot(merged_data_filtered, aes(x = Year, y = `Number.of.immigrants.in.Canada.from.2000.to.2023`)) +
+  geom_line(color = "blue") +
+  labs(title = "Number of New Immigrants in Canada from 2000-2023",
+       x = "Year",
+       y = "Number of Immigrants (Thousands)") +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5)) +  # Center the title
+  scale_y_continuous(labels = scales::comma_format(scale = 1e-3), limits = c(0, NA))  # Format y-axis labels
+
+```
+
+```{r}
+#| label: fig-hpi-plot
+#| echo: false
+#| fig-cap: House Price Index 2000-2023
 #| warning: false
@@
 ggplot(merged_data, aes(x = Year, y = Average_House_Only)) +
-  geom_line() +
-  labs(title = "Housing Price Index Over Time",
+  geom_line(color = "blue") +  # Set the line color to blue
+  labs(title = "HPI",
        x = "Year",
        y = "Average House Price Index") +
-  theme_minimal()
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5))  # Center the title
 
@@
 
-Talk more about it.
+```{r}
+#| label: fig-housing-plot
+#| fig-cap: Housing Starts Over 2000-2023
+#| echo: false
+#| warning: false
+#| message: false
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+# Find the maximum value in the 'starts' column of your dataframe
+max_starts <- max(housing_starts$Starts)
+
+# Round up the maximum value to the nearest thousand
+max_starts_rounded <- ceiling(max_starts / 1000) * 1000
+
+# Plot housing starts over time with adjusted y-axis scale
+plot(housing_starts$Year, housing_starts$Starts, type = "l", 
+     xlab = "Year", ylab = "Housing Starts", 
+     main = "Housing Starts Over Time",
+     ylim = c(0, max_starts_rounded),
+     yaxp = c(0, max_starts_rounded/1000, 10),  # Adjusting the number of ticks on the y-axis
+     col = "blue")  # Set the line color to blue
+
+# Customize the y-axis labels to display in numeric format with thousands separator
+axis(side = 2, at = axTicks(2), labels = format(axTicks(2), big.mark = ",", scientific = FALSE))
+
+```
 
 ```{r}
-#| label: immigration-plot
+#| label: hpivs-plot
 #| fig-cap: Number of immigrants in Canada from 2000 to 2023
@@
 #| message: false
-# Create a line plot for the number of immigrants in Canada over time
-ggplot(merged_data, aes(x = Year, y = `Number.of.immigrants.in.Canada.from.2000.to.2023`)) +
-  geom_line(color = "blue") +
-  labs(title = "Number of Immigrants in Canada from 2000 to 2023",
-       x = "Year",
-       y = "Number of Immigrants") +
-  theme_minimal()
+# Scatterplot of HPI against each economic indicator
+library(ggplot2)
+
+ggplot(data = merged_data, aes(x = Interest_Rate, y = Inflation.Rate....)) +
+  geom_point() + 
+  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Adding a linear trend line
+  labs(title = "Scatterplot of HPI vs Interest Rate",
+       x = "Interest Rate",
+       y = "HPI")
 
@@
 
-Talk way more about it.
-
-# Model
-
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-## Model set-up
-
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
-
-# Results
-
-Our results are summarized in @tbl-modelresults.
+# Model{#sec-data2}
 
 ```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
 
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
 ```
 
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
+### Model set-up
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
+The goal of our modeling strategy is to analyze the relationship between the economic predictors and the housing price index (HPI), and explore how the economic variables contribute to the housing market.
+After exploring some of the intial data we observed a general trend between different variables, that indicated some level of relationship between them.
+To see the actual relationships, we implement a multiple linear regression model to take a close look at what variables seem to have an effect on the outcome variable, denoted by the house price index.
 
-# Discussion
+### Multiple Linear Regression
+We chose to explore a variety of economic and count variables like immigration and housing starts, to see how these impact the housing prices we use multiple linear regression model.
 
-## First discussion point {#sec-first-point}
+ The basic form of the regression equation is:
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+The linear regression equation for predicting the House Price Index is given by:
 
-## Second discussion point
+House_Price_Index = β_0 + β_1 × Inflation_Rate + β_2 × Interest_Rate + β_3 × Unemployment_Rate + β_4 × Immigration_numbers + β_5 × Housing_Starts + ε
 
-## Third discussion point
+Where:
+\begin{itemize}
+  \item \( \text{House\_Price\_Index} \) is the response variable (dependent variable).
+  \item \( \beta_0 \) is the intercept term.
+  \item \( \beta_1, \beta_2, \beta_3, \beta_4, \beta_5 \) are the coefficients (regression coefficients) corresponding to each predictor variable.
+  \item \( \text{Inflation\_Rate}, \text{Interest\_Rate}, \text{Unemployment\_Rate}, \text{Immigration\_numbers}, \text{Housing\_Starts} \) are the predictor variables (independent variables).
+  \item \( \varepsilon \) represents the error term.
+\end{itemize}
 
-## Weaknesses and next steps
+### Model Justifications
+The above factors were chosen as they are economic indicators of the housing market, and are often the variables that have influence in the housing market given the market conditions.
+For inflation and interest rates, there is a cyclic relationship, as inflation goes up, the interest rates on mortgages go up as a direct consequence.
+This in turn usually increases monthly housing costs, detering buyers from qualifying for loan approvals.
+Unemployment rates are correlated to small spending.
+Given Canada's housing crisis, immigration numbers and housing starts seem to be changing supply and demand, as a result these indicators can be conclusive.
 
-Weaknesses and next steps should also be included.
+We expect a negative linear relationship between interest rates and housing price index, housing starts and house price index, and unemployment rate and house price index.
+For Immigration we expect a somewhat positive linear relationship to house price index.
+The increasing pace of immigration has been fueling demand and constricting the market.
+These relationships are estimated based upon economic patterns and past evidence.
 
-\newpage
+## Results {#sec-data3}
 
-\appendix
+### Model Summary
 
-# Appendix {-}
+Our multiple linear regression model is summarized in [@tbl-model-sum].
+Our multiple linear regression model reveals the strength and association of each indicator to the outcome variable, while holding other variables constant.
+The estimated coefficients in the model indicate the magnitude and direction of each variable's association.
+Indicators that had a positive relationship include the inflation_rate where each unit of increase in its value correlated to a 5.398 unit (estimate coefficient) increase in house_price_index.
 
-# Additional data details
+The variables interest rate and unemployment_rate had a negative association to the house_price_index, as expected.For each unit increase of the interest rate (percent) the house price index saw a 12.45 unit decrease.
+The unemployment rate had estimate coefficient -2.812 indicating that as the unemployment rate changed by a unit the house price index saw a 2.812 unit decrease.
+However, here the relationship is not statistically significant as p > 0.05.
 
-# Model details {#sec-model-details}
+The housing_starts suggests a small negative effect on the house price index, but is not statistically significant.
+Similarly, the immigration_numbers saw a small positive effect on the house price index, but it is not statistically significant.
+Considering the nature of these variables, they are count values, hence a change in one unit (one housing start increase or one immigrant number increase) is not expressed significantly by this model.
 
-## Posterior predictive check
+The model demonstrates an overall good fit with an R-squared value of 0.8522, indicating that approximately 85.229% of the variance in house price index is explained by the included predictors.
+The adjusted R-squared, which accounts for the number of predictors in the model, is slightly lower at 0.8088, suggesting the model is fairly robust.
+The residual standard error is 8.364, suggesting the amount of variability of observed responses around the fitted regression line (plot).
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+# Discussion{#sec-data4}
 
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+## Weaknesses and Biases 
+The data sets used for the analysis were combined and explored, these data sets may have added certain biases to our overall findings.
+There are a number of data sets that had missing values, and most of the data had to be modified to be compatible for use against the other indicators, we also note the  differences in their units reported.
+The modelling process is prone to adding bias to our findings due to model pre-assumptions, and whether the model may have overlooked trends due to its limitations.
+The regression model suggested immigration numbers and housing starts as not significant indicators of the house price index.
+However, from a economic perspective these variables should have affected the price index as they represent supply and demand variables.
 
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+Since Our model could not grasp the nature of its data and showed a very minute relationship.
+To show some key insights, transforming these variables to a concise unit of change could help highlight their importance, or additionally using already aggregated data could help the regression model uncover associations more meaningfully.
 
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-```
-
-\newpage
+## Future Research
+The analysis we covered only explores a handful of housing market predictors, in the future expanding the preidctors and ading a variety of variable types can help build more intuitive models for udenrstanding.
+To gain a new perspective on the housing market, indicatros relation t demographics, socieconmic factors, how many buyers in the market can comfortably afford mortgages and loans.
+This would build a foundation for a deeper analysis, and an opportunity to learn more about the housing market and whether Canada is headed in the right direction in the coming years.
 

### 2024-04-18 paper/paper.qmd
added model files git comm

@@
 bibliography: references.bib
+toc: true
 ---
@@
 library(scales)
+library(dplyr)
+library(knitr)
+library(kableExtra)
 
@@
 With the complexity of the housing market influenced by numerous factors, keeping track of its evolution proves challenging.
-This analysis aims to understand the dynamics of the housing market, exploring the impact of economic variables and identifying significant contributors.
+The estimand of this paper is to find the associations the of economic variables, the supply demand variables represented by housing starts and immigration numbers to the house price index.
 By examining economic indicator changes, we seek to uncover their significance and understand correlations among variables and their market impact.
@@
 The yearly immigration numbers were obtained from Statista [@immigrants], and was originally sourced from Statistics Canada.
-This data reports the number of immigrants Canada welcomes in each year from 2000-2023.
+This data reports the number of immigrants Canada welcomes in each year from 2000-2023 in thousands.
 The data was collected and sourced from a survey done by Statistics Canada.
@@
 
+```{r}
+# Load necessary libraries
+library(dplyr)
+library(knitr)
+library(kableExtra)
+
+# Display the first 6 rows of the merged data as a table using kable with proper styling
+kable(head(merged_data, 6), format = "latex", caption = "Merged Data Table") %>%
+  kable_styling(latex_options = "scale_down") %>%
+  as.character() %>%
+  cat("\n\\centering")  # Add LaTeX centering command
+
+```
 ## Data Cleaning 
@@
 We used R [@citeR] for data cleaning and processing, using packages like tidyverse [@tidy] for data manipulation and janitor [@jan] for cleaning column names.
-Other packages used includes `ggplot2` [@ggplot], `dplyr` [@dp], `readr` [@read], `tibble` [@tib], `janitor` [@jan], `knitr` [@knit], `ggbeeswarm` [@gg], `ggrepel` [@repel], `kableExtra`[@kable], `readxl`[@readxl], `rstanarm`[@rstan], `modelsummary`[@model] and `here` [@here].
+Other packages used includes `ggplot2` [@ggplot], `dplyr` [@dp], `readr` [@read], `tibble` [@tib], `janitor` [@jan], `knitr` [@knit], `ggrepel` [@repel], `kableExtra`[@kable], `readxl`[@readxl], `rstanarm`[@rstan], `modelsummary`[@model] and `here` [@here].
 All these different data sets were cleaned in a different way to match the final representation of the cleaned data.
@@
 ## Data Modifications
-Overall, the values were often reported at different times, for consistency we chose to use a yearly metric. We averaged upon quarterly or monthly reported data to gain an averaged yearly value.
+Overall, the values were often reported at different times, for consistency we chose to use a yearly metric.
+We averaged upon quarterly or monthly reported data to gain an averaged yearly value.
+We then created a merged datadframe joined by the year variable and achieved our cleaned data as seen in @tbl-clean-data.
 
@@
 ```{r}
-#| label: hpivs-plot
+#| label: fig-hpivs-plot
 #| fig-cap: Number of immigrants in Canada from 2000 to 2023
@@
 
-House_Price_Index = β_0 + β_1 × Inflation_Rate + β_2 × Interest_Rate + β_3 × Unemployment_Rate + β_4 × Immigration_numbers + β_5 × Housing_Starts + ε
+\[ \text{House\_Price\_Index} = \beta_0 + \beta_1 \times \text{Inflation\_Rate} + \beta_2 \times \text{Interest\_Rate} + \beta_3 \times \text{Unemployment\_Rate} + \beta_4 \times \text{Immigration\_numbers} + \beta_5 \times \text{Housing\_Starts} + \epsilon \]
 
@@
 
-## Results {#sec-data3}
+# Results {#sec-data3}
 
@@
 
+## Interest Rates and HPI
+The model suggests interest rate as a significant predictor of HPI
+
 ## Weaknesses and Biases 
@@
 
-Since Our model could not grasp the nature of its data and showed a very minute relationship.
+Since our model could not grasp the nature of its data and showed a very minute relationship.
 To show some key insights, transforming these variables to a concise unit of change could help highlight their importance, or additionally using already aggregated data could help the regression model uncover associations more meaningfully.

### 2024-04-18 scripts/02-data_cleaning.R

@@ -9,6 +9,8 @@
 #### Workspace setup ####
 library(tidyverse)
 library(dplyr)
+install.packages("readxl")  # Install the readxl package if not already installed
+library(readxl)  # Load the readxl package
 
 #### Clean data ####
 inflation_data <- read_csv("data/raw_data/inflation-rate-cpi.csv", skip = 16)
@@ -16,6 +18,7 @@ unemployment_data <- read_csv("data/raw_data/canada-unemployment-rate.csv", skip
 interest_data <- read_excel("data/raw_data/interest_rate_2000.xlsx",skip=, col_types = "text")
 hpi_index <- read_excel("data/raw_data/HPI_20-23.xlsx",skip=, col_types = "text")
 immigration <- read_excel("data/raw_data/immigration_data.xlsx",skip=, col_types = "text")
+housing_starts <- read_excel("data/raw_data/housing_starts.xlsx",skip=, col_types = "text")
 # inflation
 # Clean data
 inflation_data_cleaned <- inflation_data %>%
@@ -154,8 +157,6 @@ immigration <- rbind(immigration, new_row)
 immigration$`Number of immigrants in Canada from 2000 to 2023` <- as.numeric(immigration$`Number of immigrants in Canada from 2000 to 2023`)
 
 # Calculate percent change
-immigration$`Percent Change` <- c(NA, diff(as.numeric(immigration$`Number of immigrants in Canada from 2000 to 2023`)) / lag(as.numeric(immigration$`Number of immigrants in Canada from 2000 to 2023`)) * 100)
-
 # Print the modified dataframe
 print(tail(immigration))
 
@@ -174,33 +175,84 @@ print(immigration)
 
 
 
-# Merge the cleaned datasets by the "year" property
-merged_data <- merge(inflation_data_cleaned, unemployment_cleaned, by = "year", all = TRUE)
+# cleaning housing starts 
+print(housing_starts)
 
+
+# Find the row indices where "housing starts" and "housing estimates" are located in the first column
+housing_starts_row <- which(housing_starts[, 1] == "Housing starts")
+housing_estimates_row <- which(housing_starts[, 1] == "Housing estimates")
+
+# Extract the rows based on the found indices
+housing_starts_data <- housing_starts[housing_starts_row, ]
+housing_estimates_data <- housing_starts[housing_estimates_row, ]
+
+# Combine the two dataframes
+combined_data <- rbind(housing_starts_data, housing_estimates_data)
+print(combined_data)
[…]

### 2024-04-18 scripts/04-model.R

@@ -1,12 +1,9 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Model data for meaningful analysis
+# Author: Navya Hooda
+# Date: April 9, 2023
+# Contact: navya.hooda@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
 
 #### Workspace setup ####
 library(tidyverse)

### 2024-04-18 scripts/04-model.R
added model

@@ -8,27 +8,39 @@
 #### Workspace setup ####
 library(tidyverse)
 library(rstanarm)
-
+library(modelsummary)
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+merged_data <- read_csv("data/analysis_data/merged_data.csv")
+housing_starts <- read_csv("data/analysis_data/total_starts_by_year.csv")
+#
+print(merged_data)
+# Assuming merged_data is your merged dataset
+# Assuming housing_starts is your dataset containing the number of housing starts
 
-### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
+
+# Subset the relevant columns from merged_data
+model_data <- merged_data[, c("Average_House_Only", "Inflation Rate (%)", "Interest_Rate", "Unemployment Rate (%)", "Number of immigrants in Canada from 2000 to 2023", "Starts")]
+
+# Rename columns for convenience
+colnames(model_data) <- c("House_Price_Index", "Inflation_Rate", "Interest_Rate", "Unemployment_Rate", "Immigration_numbers", "Housing_Starts")
+print(colnames(model_data))
+# Assuming 'merged_data' is your merged dataset containing the housing starts data
+
+# Fit a linear regression model with housing starts as an additional predictor
+pricing_model <- lm(House_Price_Index ~ Inflation_Rate + Interest_Rate + Unemployment_Rate + Immigration_numbers + Housing_Starts, data = model_data)
+
+# Summary of the model
+summary(pricing_model)
 
 
 #### Save model ####
 saveRDS(
-  first_model,
-  file = "models/first_model.rds"
+  pricing_model,
+  file = "models/pricing_model.rds"
 )
 
+saveRDS(
+  lm_model,                  # Your linear regression model object
+  file = "models/lm_model.rds"  # File path where you want to save the model
+)
 

### 2024-04-19 paper/paper.qmd
added remaining scripts, final paper changes, deleted files

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper examines Ontario's housing market over 20 years, focusing on economic variables, housing starts, and immigration's impact on house prices.
+We evaluate the current housing development initiatives in addressing the housing crisis, along with immigration targets as supply and demand variables.
+The market's current state indicates an already constricted inventory that is largely influenced by interest rate hikes over the last few years."
+header-includes:
+  - \usepackage{amsmath}
+  - \usepackage{amssymb}
 format: pdf 
@@
 
-Focusing particularly on Ontario's housing landscape, where activity has surged in recent years, this paper investigates shifts in housing demand over the past two decades.
+Focusing particularly on Ontario's housing market, where activity has surged in recent years, this paper investigates shifts in housing demand over the past two decades.
 We evaluate whether current housing development initiatives adequately address the housing crisis.
@@
 Through this analysis, we aim to grasp the current reality of the housing market and understand how economic conditions, including factors like immigration and housing starts, influence its trajectory.
-With the housing market facing constraints such as high interest rates, we seek to uncover insights regarding the future of the market.
+With the housing market facing constraints such as high interest rates, we found that they did have a significant correlation to changes in house price index, whereas certain factors did not have much association to house price index at all.
 
-The remainder of this paper is structured as follows.In the Data@sec-data, we note the sources of the data sets utilized in the paper, the specific techniques we used to process them, and key variables we wish to use.
+The remainder of this paper is structured as follows.In the Data @sec-data, we note the sources of the data sets utilized in the paper, the specific techniques we used to process them, and key variables we wish to use.
 In the Model @sec-data2, we created models to assess the impact of the key variables on our outcome of housing prices.
@@
 ```{r} 
-# Load necessary libraries
-library(dplyr)
-library(knitr)
+#| label: tbl-cleaned-data
+#| message: false
+#| echo: false
+#| warning: false
+#| tbl-cap: "Clean and merged data"
 library(kableExtra)
 
-# Display the first 6 rows of the merged data as a table using kable with proper styling
-kable(head(merged_data, 6), format = "latex", caption = "Merged Data Table") %>%
-  kable_styling(latex_options = "scale_down") %>%
-  as.character() %>%
-  cat("\n\\centering")  # Add LaTeX centering command
+options(kableExtra.latex.max.width=1)
+head(merged_data, 6) |>
+  kable(
+        col.names = c("Year","Average_House_Only","Interest_Rate","Unemployment Rate (%)","Annual Change.x","Inflation Rate (%)","Annual Change.y","Number of immigrants in Canada from 2000 to 2023","Starts"),
+    booktabs = TRUE,
+    longtable = FALSE
+  ) %>%
+  kable_styling(latex_options = c("striped", "scale_down"), font_size = 3)
 
@@
 All these different data sets were cleaned in a different way to match the final representation of the cleaned data.
+
+```{r}
+#| label: tbl-model-sum
+#| fig-cap: "Model Summary"
+#| echo: false
+#| warning: false
+#| message: false
+
+# Load necessary libraries
+library(dplyr)
+library(knitr)
+library(kableExtra)
+library(here)
+
+# Load the model file
+pricing_model <-
+  readRDS(file = here::here("models/pricing_model.rds"))
+
+# # Now you can access model_data
+# # For example, if model_data is a data frame in the model file
+# 
+# 
+# # Create a function to format the coefficients table
+# format_coefficients_table <- function(model_summary) {
+#   # Extract coefficients, standard errors, t-values, and p-values
+#   coefficients <- coef(model_summary)
+#   std_errors <- summary(model_summary)$coefficients[, "Std. Error"]
+#   t_values <- summary(model_summary)$coefficients[, "t value"]
+#   p_values <- summary(model_summary)$coefficients[, "Pr(>|t|)"]
+#   
+#   # Create a data frame
+#   coefficients_table <- data.frame(
+#     Coefficient = coefficients,
+#     `Std. Error` = std_errors,
+#     `t value` = t_values,
+#     `Pr(>|t|)` = p_values
+#   )
+#   
+#   return(coefficients_table)
+# }
+# 
+# # Call the function with your model summary
+# model_summary <- lm(House_Price_Index ~ Inflation_Rate + Interest_Rate + 
+#                       Unemployment_Rate + Immigration_numbers + Housing_Starts, 
+#                     data = model_data)
+# coefficients_table <- format_coefficients_table(model_summary)
+# 
+# # Print the coefficients table using kable and add formatting
+# kable(coefficients_table, format = "markdown", align = "c") %>%
+#   kable_styling(latex_options = c("striped", "scale_down")) %>%
+#   add_header_above(c(" " = 1, "Coefficients" = 4))
+
+```
+
+```{r}
+#| label: tbl-model
+#| tbl-cap: Model Summary Results 
+#| echo: false
+#| warning: false
+#| message: false
+modelsummary::modelsummary(
+  list(
+    "Price model" = pricing_model
+  ),
+  fmt = 2
+)
+
+```
 
@@
 We averaged upon quarterly or monthly reported data to gain an averaged yearly value.
-We then created a merged datadframe joined by the year variable and achieved our cleaned data as seen in @tbl-clean-data.
-
-```{r}
-# Load the necessary packages
-#| label: fig-immigration-plot
-#| fig-cap: Number of immigrants in Canada from 2000 to 2023
-#| echo: false
-#| warning: false
-#| message: false
-# Filter out rows with missing values in the 'Number.of.immigrants.in.Canada.from.2000.to.2023' column
-merged_data_filtered <- merged_data[!is.na(merged_data$Number.of.immigrants.in.Canada.from.2000.to.2023),]
-
-# Create a line plot for the number of immigrants in Canada over time
-ggplot(merged_data_filtered, aes(x = Year, y = `Number.of.immigrants.in.Canada.from.2000.to.2023`)) +
-  geom_line(color = "blue") +
-  labs(title = "Number of New Immigrants in Canada from 2000-2023",
-       x = "Year",
-       y = "Number of Immigrants (Thousands)") +
-  theme_minimal() +
-  theme(plot.title = element_text(hjust = 0.5)) +  # Center the title
-  scale_y_continuous(labels = scales::comma_format(scale = 1e-3), limits = c(0, NA))  # Format y-axis labels
-
-```
+We then created a merged datadframe joined by the year variable and achieved our cleaned data as seen in @tbl-cleaned-data.
 
@@
 #| label: fig-housing-plot
-#| fig-cap: Housing Starts Over 2000-2023
+#| fig-cap: Housing Starts Over 2000-2023 in Ontario
 #| echo: false
@@
      xlab = "Year", ylab = "Housing Starts", 
-     main = "Housing Starts Over Time",
+     main = "",
      ylim = c(0, max_starts_rounded),
@@
 axis(side = 2, at = axTicks(2), labels = format(axTicks(2), big.mark = ",", scientific = FALSE))
-
-```
-
-```{r}
-#| label: fig-hpivs-plot
-#| fig-cap: Number of immigrants in Canada from 2000 to 2023
-#| echo: false
-#| warning: false
-#| message: false
-# Scatterplot of HPI against each economic indicator
-library(ggplot2)
-
-ggplot(data = merged_data, aes(x = Interest_Rate, y = Inflation.Rate....)) +
-  geom_point() + 
-  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Adding a linear trend line
-  labs(title = "Scatterplot of HPI vs Interest Rate",
-       x = "Interest Rate",
-       y = "HPI")
 
@@
 The goal of our modeling strategy is to analyze the relationship between the economic predictors and the housing price index (HPI), and explore how the economic variables contribute to the housing market.
-After exploring some of the intial data we observed a general trend between different variables, that indicated some level of relationship between them.
-To see the actual relationships, we implement a multiple linear regression model to take a close look at what variables seem to have an effect on the outcome variable, denoted by the house price index.
+After exploring some of the initial data we observed a general trend between different variables, that indicated some level of relationship between them.
+To see the actual relationships, we implement a bayesian generalized linear model to take a close look at what variables seem to have an effect on the outcome variable, denoted by the house price index.
 
-### Multiple Linear Regression
-We chose to explore a variety of economic and count variables like immigration and housing starts, to see how these impact the housing prices we use multiple linear regression model.
+###  Bayesian linear regression
+We chose to explore a variety of economic and count variables like immigration and housing starts, to see how these impact the housing prices we use a Bayesian GLM.
 
@@
 
-\[ \text{House\_Price\_Index} = \beta_0 + \beta_1 \times \text{Inflation\_Rate} + \beta_2 \times \text{Interest\_Rate} + \beta_3 \times \text{Unemployment\_Rate} + \beta_4 \times \text{Immigration\_numbers} + \beta_5 \times \text{Housing\_Starts} + \epsilon \]
+$$
+\begin{aligned}
+\text{House\_Price\_Index} = & \beta_0 + \beta_1 \times \text{Inflation\_Rate} \\
+& + \beta_2 \times \text{Interest\_Rate} + \beta_3 \times \text{Unemployment\_Rate} \\
+& + \beta_4 \times \text{Immigration\_numbers} + \beta_5 \times \text{Housing\_Starts} + \epsilon
+\end{aligned}
+$$
 
@@
 
-Our multiple linear regression model is summarized in [@tbl-model-sum].
+Our multiple linear regression model is summarized in [@tbl-model].
 Our multiple linear regression model reveals the strength and association of each indicator to the outcome variable, while holding other variables constant.
 The estimated coefficients in the model indicate the magnitude and direction of each variable's association.
-Indicators that had a positive relationship include the inflation_rate where each unit of increase in its value correlated to a 5.398 unit (estimate coefficient) increase in house_price_index.
+Indicators that had a positive relationship include the inflation_rate where each unit of increase in its value correlated to a 3.046 unit (estimate coefficient) increase in house_price_index.
 
-The variables interest rate and unemployment_rate had a negative association to the house_price_index, as expected.For each unit increase of the interest rate (percent) the house price index saw a 12.45 unit decrease.
-The unemployment rate had estimate coefficient -2.812 indicating that as the unemployment rate changed by a unit the house price index saw a 2.812 unit decrease.
-However, here the relationship is not statistically significant as p > 0.05.
+The variables interest rate and unemployment_rate had a negative association to the house_price_index, as expected.For each unit increase of the interest rate (percent) the house price index saw a 11.579 unit decrease.
+The unemployment rate had estimate coefficient -2.320 indicating that as the unemployment rate changed by a unit the house price index saw a 2.320 unit decrease.
+However, here the relationship is not statistically significant.
 
@@
 
-The model demonstrates an overall good fit with an R-squared value of 0.8522, indicating that approximately 85.229% of the variance in house price index is explained by the included predictors.
-The adjusted R-squared, which accounts for the number of predictors in the model, is slightly lower at 0.8088, suggesting the model is fairly robust.
-The residual standard error is 8.364, suggesting the amount of variability of observed responses around the fitted regression line (plot).
+The model demonstrates an overall good fit with an R-squared value of 0.842, indicating that approximately 84.2% of the variance in house price index is explained by the included predictors.
+The adjusted R-squared, which accounts for the number of predictors in the model, is slightly lower at 0.745, suggesting the model is fairly robust.
+The log likelihood has a value of -74.00 and indicates that the model provides a relatively good fit, as this refers to how well the model explains the observed data.
 
@@
 ## Interest Rates and HPI
-The model suggests interest rate as a significant predictor of HPI
+The model suggests interest rate as a significant predictor of HPI, or one that has the most fluctuation in a one unit change to our response variable of house price index.
+This trend was expected and is justified considering economic tendencies.
+The interest rates in our study recently saw a consistent hike over the last few years.
+This interest rate is the measure of the mortgage lending rate, a variable directly correlated to housing affordability and prices.
+As these have gone up, this negative relationship towards the house price index is justified.
+In terms of the housing market, interest rates show to be beneficial for generally decreasing house price index, and likely other market predictors not evaluated in this study.
 
@@
 
+The
+
 ## Future Research

### 2024-04-19 scripts/02-data_cleaning.R

@@ -11,7 +11,7 @@ library(tidyverse)
 library(dplyr)
 install.packages("readxl")  # Install the readxl package if not already installed
 library(readxl)  # Load the readxl package
-
+library(arrow)
 #### Clean data ####
 inflation_data <- read_csv("data/raw_data/inflation-rate-cpi.csv", skip = 16)
 unemployment_data <- read_csv("data/raw_data/canada-unemployment-rate.csv", skip = 16)
@@ -143,7 +143,7 @@ immigration$Year <- sub(".* - ", "", immigration$Year)
 print(immigration)
 
 # Create a new row for 2000 with NA immigration value
-new_row <- data.frame(Year = "2000", `Number of immigrants in Canada from 2000 to 2023` = 0)
+new_row <- data.frame(Year = "2000", `Number of immigrants in Canada from 2000 to 2023` = 252527)
 
 # Ensure column names match
 colnames(new_row) <- colnames(immigration)
@@ -303,7 +303,7 @@ print(duplicated_rows)
 
 # Save the merged data to a CSV file
 write.csv(merged_data, "data/analysis_data/merged_data.csv", row.names = FALSE)
-
+write_parquet(merged_data, "data/analysis_data/merged.parquet")
 
 # Print the merged data
 print(merged_data)

### 2024-04-19 scripts/03-test_data.R

@@ -1,15 +1,52 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Test data for checks 
+# Author: Navya Hooda
+# Date: April 9, 2023
+# Contact: navya.hooda@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+
+
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(testthat)
+
 
 #### Test data ####
+merged_data <- read.csv("data/analysis_data/merged_data.csv")
+
+
+
+#### Test data ####
+
+# Test the structure and type of the economic variables
+test_that("Economic variables columns are of the correct type", {
+  expect_type(merged_data$Interest_Rate, "double")
+  expect_type(merged_data$Inflation_Rate, "numeric")
+  expect_type(merged_data$Unemployment_Rate, "numeric")
+})
+
+# Test the structure and type of the housing starts data
+test_that("Housing starts data columns are of the correct type", {
+  expect_type(merged_data$Year, "integer")
+  expect_type(merged_data$Housing_Starts, "numeric")
+})
+
+# Test the structure and type of the immigration data
+test_that("Immigration data columns are of the correct type", {
+  expect_type(merged_data$Year, "integer")
+  expect_type(merged_data$Immigrants, "numeric")
+})
+
+# Test that the years are within the expected range
+test_that("Year values are within the expected range", {
+  expect_true(all(merged_data$Year >= 2000 & merged_data$Year <= 2024))
+})
+
+# Test for missing values
+test_that("There are no missing values in key columns", {
+  expect_true(all(!is.na(merged_data$Interest_Rate)))
+  expect_true(all(!is.na(merged_data$Housing_Starts)))
+  expect_true(all(!is.na(merged_data$Immigrants)))
+})

### 2024-04-19 scripts/04-model.R

@@ -27,20 +27,17 @@ print(colnames(model_data))
 # Assuming 'merged_data' is your merged dataset containing the housing starts data
 
 # Fit a linear regression model with housing starts as an additional predictor
-pricing_model <- lm(House_Price_Index ~ Inflation_Rate + Interest_Rate + Unemployment_Rate + Immigration_numbers + Housing_Starts, data = model_data)
+pricing_model <- stan_glm(House_Price_Index ~ Inflation_Rate + Interest_Rate + Unemployment_Rate + Immigration_numbers + Housing_Starts, data = model_data)
 
 # Summary of the model
 summary(pricing_model)
+modelsummary(pricing_model)
+
 
 
 #### Save model ####
 saveRDS(
   pricing_model,
   file = "models/pricing_model.rds"
-)
-
-saveRDS(
-  lm_model,                  # Your linear regression model object
-  file = "models/lm_model.rds"  # File path where you want to save the model
 )
 

### 2024-04-19 paper/paper.qmd
final changes

@@
 We used R [@citeR] for data cleaning and processing, using packages like tidyverse [@tidy] for data manipulation and janitor [@jan] for cleaning column names.
-Other packages used includes `ggplot2` [@ggplot], `dplyr` [@dp], `readr` [@read], `tibble` [@tib], `janitor` [@jan], `knitr` [@knit], `ggrepel` [@repel], `kableExtra`[@kable], `readxl`[@readxl], `rstanarm`[@rstan], `modelsummary`[@model] and `here` [@here].
+Other packages used includes `ggplot2` [@ggplot], `dplyr` [@dp], `tibble` [@tib], `janitor` [@jan], `knitr` [@knit], `ggrepel` [@repel], `kableExtra`[@kable], `readxl`[@readxl], `rstanarm`[@rstan], `modelsummary`[@model] and `here` [@here].
 All these different data sets were cleaned in a different way to match the final representation of the cleaned data.
@@
 After exploring some of the initial data we observed a general trend between different variables, that indicated some level of relationship between them.
-To see the actual relationships, we implement a bayesian generalized linear model to take a close look at what variables seem to have an effect on the outcome variable, denoted by the house price index.
+To see the actual relationships, we implement a Bayesian generalized linear model to take a close look at what variables seem to have an effect on the outcome variable, denoted by the house price index.
 
-###  Bayesian linear regression
+###  Bayesian Linear Regression 
 We chose to explore a variety of economic and count variables like immigration and housing starts, to see how these impact the housing prices we use a Bayesian GLM.
@@
 
-The linear regression equation for predicting the House Price Index is given by:
+The model equation for predicting the House Price Index is given by:
 
@@
 
-Our multiple linear regression model is summarized in [@tbl-model].
-Our multiple linear regression model reveals the strength and association of each indicator to the outcome variable, while holding other variables constant.
+Our Bayesian linear regression model is summarized in [@tbl-model].
+Our model reveals the strength and association of each indicator to the outcome variable, while holding other variables constant.
 The estimated coefficients in the model indicate the magnitude and direction of each variable's association.

### 2024-04-19 scripts/02-data_cleaning.R

@@ -303,7 +303,7 @@ print(duplicated_rows)
 
 # Save the merged data to a CSV file
 write.csv(merged_data, "data/analysis_data/merged_data.csv", row.names = FALSE)
-write_parquet(merged_data, "data/analysis_data/merged.parquet")
+write_parquet(merged_data, "data/analysis_data/merged_data.parquet")
 
 # Print the merged data
 print(merged_data)

### 2024-04-19 scripts/04-model.R

@@ -9,6 +9,8 @@
 library(tidyverse)
 library(rstanarm)
 library(modelsummary)
+
+
 #### Read data ####
 merged_data <- read_csv("data/analysis_data/merged_data.csv")
 housing_starts <- read_csv("data/analysis_data/total_starts_by_year.csv")

### 2024-04-19 paper/paper.qmd
minor changes to abstract

@@
 abstract: "This paper examines Ontario's housing market over 20 years, focusing on economic variables, housing starts, and immigration's impact on house prices.
-We evaluate the current housing development initiatives in addressing the housing crisis, along with immigration targets as supply and demand variables.
-The market's current state indicates an already constricted inventory that is largely influenced by interest rate hikes over the last few years."
+We evaluate the current housing development initiatives in addressing the housing crisis, along with immigration targets as supply and demand variables.The paper finds an already constricted market that is largely influenced by interest rate hikes during a housing crisis."
 header-includes:

### 2024-04-19 paper/paper.qmd
final changes x 2
added parquet files, final changed x 3

@@
 
-With the complexity of the housing market influenced by numerous factors, keeping track of its evolution proves challenging.
-The estimand of this paper is to find the associations the of economic variables, the supply demand variables represented by housing starts and immigration numbers to the house price index.
-By examining economic indicator changes, we seek to uncover their significance and understand correlations among variables and their market impact.
-
 Focusing particularly on Ontario's housing market, where activity has surged in recent years, this paper investigates shifts in housing demand over the past two decades.
@@
 The unemployment rate generally refers to the percentage of the labor force that is unemployed and actively seeking employment.
-A visual of the rates discussed so far is seen in @fig-int-inf-plot.
 
@@
 
-## Weaknesses and Biases 
+## Housing Starts, Immigration Numbers and HPI
+The housing starts variable did not show any significant correlation to house price index, as it was not statistically significant.
+However, housing starts are representative of the supply in the housing market hence it should have shown some correlation to the pricing index.
+It's possible our model did not capture the correlation, but a reason for this could be that the number of housing starts has no effect on pricing because we add just the correct amount to keep prices balanced.
+However, considering Canadian regions are in a housing crisis, this is highly unlikely.
+
+If we consider the immigration as the demand factor, compared against that our housing price index we expected some effect on the pricing.
+As the demand or immigrants increase, the housing prices should have had some proportionally increase.
+Especially, since Canada has increased immigration by 20.3% during 2022-2023 and will continue to increase each year [@reuters_canada_population].
+Given that, the numbers suggest the demand has been growing rapidly, it is likely the model was not able to capture immigration as a significant factor to the housing market due to model limitations.
+
+## Biases and Limitations
 The data sets used for the analysis were combined and explored, these data sets may have added certain biases to our overall findings.
@@
 
-The
+Additionally, our choice of housing market indicator can be a concern when generalizing our trends to the overall Canadian housing market.
+We chose to assign the house price index, specifically representing the house price index of new homes as our market indicator, this predictor does not account for pricing index of older homes.
+The market inOntario is at a pace where buyers are equally interested in older homes as opposed to new homes due to the limited supply of housing that has arisen in the past years.
+By only analyzing the new house price index, we may have introduced some limitations as to how representative our model's results are of the housing market as a whole.
 
 ## Future Research
-The analysis we covered only explores a handful of housing market predictors, in the future expanding the preidctors and ading a variety of variable types can help build more intuitive models for udenrstanding.
-To gain a new perspective on the housing market, indicatros relation t demographics, socieconmic factors, how many buyers in the market can comfortably afford mortgages and loans.
-This would build a foundation for a deeper analysis, and an opportunity to learn more about the housing market and whether Canada is headed in the right direction in the coming years.
+
+To gain a deeper understanding of the Canadian housing market, it would be helpful to analyze various metropolitan and growing regions across the country.
+By including additional economic factors such as GDP and average wages, we can better understand how these variables influence multiple outcomes, including average house prices.
+
+The analysis we've done so far only covers a few predictors, so it would be beneficial to expand our scope.
+Including indicators related to demographics, socioeconomic factors, and the affordability of mortgages can provide more insights.
+For example, exploring housing market characteristics, prices based on dwelling types, and the demographics of recent buyers can offer a more comprehensive view.
+
+By analyzing these factors together, we can build a stronger foundation for understanding the housing market and gaining insights into Canada's future direction.
+
+\newpage
 
